---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, eval = T, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# blockclustr

The goal of blockclustr is to allow researchers to easily construct experimental designs where they have to assign units to clusters themselves.  Usually, cluster randomized experiments are used when the clusters are naturally occurring, like cities or classrooms.  But sometimes, researchers will bring units together that aren't naturally clustered, particularly for experiments involving social interaction.

Since the researcher is already burdened with assigning units to clusters in these settings, `blockclustr` also implements *blocked* randomization.  In sum, `blockclustr` helps a research assign units to clusters in a way that increases precision of their estimated causal effects by simultaneously creating blocked clusters.  


## Installation

You can install the `blockclustr` package from Github.

```{r, eval=F}
install.packages("devtools")
devtools::install_github("erossiter/blockclustr")
```

After install, you should load the package.

```{r, eval=T}
library(blockclustr)
```

You'll also need to install the `blockTools` package.  It's not available on CRAN, so I recommend installing the latest available version (0.6-3) from the CRAN mirror on GitHub. 

```{r, eval = F}
devtools::install_github("cran/blockTools")
```


## Examples

Constructing an experimental design with `blockclustr` requires pre-treatment covariate(s).  To demonstrate the `blockclustr` functionality, I'll simulate for variables for 100 participants:

1. `id` a unique identifier for each unit
2. `pid` a string indicating partisanship (either "R" for Republican, or "D" for Democrat)
3. `inc` a numeric indicating income
4. `age` a numeric indicating age

```{r}
set.seed(123)
N <- 100
df <- data.frame(id = stringi::stri_rand_strings(N, 10),
                 pid = sample(c("R", "D"),
                              size = N,
                              replace = T),
                 inc = 20000 * rgamma(N, 1.4, 0.65),
                 age = round(runif(N, 18, 85)))

```

### Two arm design with two units per cluster

We'll first construct a design with two units per cluster, and two clusters per block.  We'll block on `age` and `income`.

```{r}
twoarm <- blockclustr(data = df,
                      n_units_per_clust = 2,
                      n_tr = 2,
                      id_var = "id",
                      block_vars = c("age", "inc"))
```

By looking at the first few rows of the design, we can see each block has two clusters, each cluster has two units, and each cluster within a block is assigned to a different experimental condition.

```{r}
head(twoarm$design, n = 8)
```


### Increasing experimental conditions and cluster size

It is simple to move beyond a two-arm design or two units per cluster by specifying the `n_tr` and `n_units_per_clust` parameters.

```{r}
multiarm <- blockclustr(data = df,
                        n_units_per_clust = 4,
                        n_tr = 3,
                        id_var = "id",
                        block_vars = c("age", "inc"))

head(multiarm$design, n = 16)
```


### Including a cluster constraint

Next we can construct a similar design, but constrain the clusters to have certain partisan makeups.  For example, below I construct a design where there are four units per cluster, but each cluster must contain a partisan majority and minority (3 Republicans and 1 Democrat or vice versa).

Note that when implementing a clustering constraint, the number of units per cluster will simply be determined by the length of the elements of the `constraint_list` parameter.  

```{r}
multiarm_cons <- blockclustr(data = df,
                             n_units_per_clust = NA,
                             n_tr = 3,
                             id_var = "id",
                             block_vars = c("age", "inc"),
                             constraint_var = "pid",
                             constraint_list = list(c("R", "R", "R", "D"),
                                                    c("D", "D", "D", "R")))

head(multiarm_cons$design, n = 24)
```
